Runtime-Assisted Convergence in Replicated Data Types

### 2022-03-21 abstract-semantics.tex
Edits for shepherd review and camera-ready

  \item $N : \texttt{Version} \rightarrow \texttt{Value}$ is a
    (partial) function (i.e., a map) that mas versions to their
    values.  We distinguish versions from values as different versions
    on different branches may store the same value (e.g., the same
    string ``hello''), yet need to be uniquely identified. The domain
    of values is left uninterpreted except for the requirement that
    it be \emph{mergeable}, i.e., define a three-way \C{merge}
    function.

@@

\paragraph{Notation} We let $v_{\odot}$ denote the initial version
(``root''), and $b_{\odot}$ denote the initial branch (``master'').
The execution of the abstract machine progresses by adding to the sets
$V$ and $E$, and updating the maps $N$, $C$, $H$, and $L$. We adopt
the usual update notation, e.g., $H[b \mapsto v]$ is a map $H'$ such
that $H'(b) = v$, and forall $b' \neq b$, $H'(b') = H(b')$.  Multiple
updates to a map are parsed left-associatively. Map $L$ is assumed to
be commutative, so $L[(b,b') \mapsto v]$ is equal to $L[(b,b') \mapsto
v][(b',b) \mapsto v]$; the former is used as a succinct replacement of
the latter. To update multiple bindings in $L$, we use the set
comprehension notation: given a branch $b$, $L[\{(b,b') \mapsto v
\,|\, \phi(b,b')\}]$ updates all bindings $(b,b')$ in $L$ to $v$,
where $b'$ is any branch such that $\phi(b,b')$ is true. Domain and
co-domain of a map $M$ is denoted $dom(M)$ and $codom(M)$
respectively.

@@

\begin{lemma}[Commit sets grow monotonically]
  \label{lem:commit-monotonicity}
  In every reachable state $\Delta = ((V,E),N,C,H,L)$ of the abstract
  machine: Forall $v_1,v_2 \in V$, if $v_1 \reaches v_2$ then $C(v_1)
  \subseteq C(v_2)$.
\end{lemma}

@@

\begin{corollary}[Commit sets modulo LCA are disjoint]
  \label{lem:commit-disjointness}
  In every reachable state $\Delta = ((V,E),N,C,H,L)$ of the abstract
  machine: Forall distinct $b_1, b_2 \in dom(H)$, and $v_0, v_1, v_2
  \in V$ s.t.  $v_1 = H(b_1)$ and $v_2 = H(b_2)$ and $v_0 =
  L(b_1,b_2)$, the following is true: $(C(v_2) - C(v_0)) ~\cap~
  (C(v_1) - C(v_0)) = \emptyset$.
\end{corollary}

@@

\begin{theorem}[{\bf Convergence}]
  \label{thm:convergence}
  In every reachable state $\Delta$ = $((V,E),N,C,H,L)$ of the abstract
  machine: Forall distinct $b_1, b_2 \in dom(H)$, and $v_1, v_2 \in V$
  such that $v_1 = H(b_1)$ and $v_2 = H(b_2)$, the following is true:
  $C(v_1) = C(v_2) \Rightarrow N(v_1) = N(v_2)$.
\end{theorem}

>>>>

  \item $N : \texttt{Version} \rightarrow \texttt{Value}$ is a
    (partial) function (i.e., a map) that maps versions to their
    values.  We distinguish versions from values as different versions
    on different branches may store the same value (e.g., the same
    string ``hello''), yet need to be uniquely identified. The domain
    of values is left uninterpreted except for the requirement that
    it be \emph{mergeable}, i.e., define a three-way \C{merge}
    function.

@@

\paragraph{Notation} We let $v_{\odot}$ denote the initial version
(``root''), and $b_{\odot}$ denote the initial branch (``master'').
The execution of the abstract machine progresses by adding to the sets
$V$ and $E$, and updating the maps $N$, $C$, $H$, and $L$. We adopt
the usual update notation, e.g., $H[b \mapsto v]$ is a map $H'$ such
that $H'(b) = v$, and for all $b' \neq b$, $H'(b') = H(b')$.  Multiple
updates to a map are parsed left-associatively. Map $L$ is assumed to
be commutative, so $L[(b,b') \mapsto v]$ is equal to $L[(b,b') \mapsto
v][(b',b) \mapsto v]$; the former is used as a succinct replacement of
the latter. To update multiple bindings in $L$, we use the set
comprehension notation: given a branch $b$, $L[\{(b,b') \mapsto v
\,|\, \phi(b,b')\}]$ updates all bindings $(b,b')$ in $L$ to $v$,
where $b'$ is any branch such that $\phi(b,b')$ is true. Domain and
co-domain of a map $M$ is denoted $dom(M)$ and $codom(M)$
respectively.

@@

\begin{lemma}[Commit sets grow monotonically]
  \label{lem:commit-monotonicity}
  In every reachable state $\Delta = ((V,E),N,C,H,L)$ of the abstract
  machine: For all $v_1,v_2 \in V$, if $v_1 \reaches v_2$ then $C(v_1)
  \subseteq C(v_2)$.
\end{lemma}

@@

\begin{corollary}[Commit sets modulo LCA are disjoint]
  \label{lem:commit-disjointness}
  In every reachable state $\Delta = ((V,E),N,C,H,L)$ of the abstract
  machine: For all distinct $b_1, b_2 \in dom(H)$, and $v_0, v_1, v_2
  \in V$ s.t.  $v_1 = H(b_1)$ and $v_2 = H(b_2)$ and $v_0 =
  L(b_1,b_2)$, the following is true: $(C(v_2) - C(v_0)) ~\cap~
  (C(v_1) - C(v_0)) = \emptyset$.
\end{corollary}

@@

\begin{theorem}[{\bf Convergence}]
  \label{thm:convergence}
  In every reachable state $\Delta$ = $((V,E),N,C,H,L)$ of the abstract
  machine: For all distinct $b_1, b_2 \in dom(H)$, and $v_1, v_2 \in V$
  such that $v_1 = H(b_1)$ and $v_2 = H(b_2)$, the following is true:
  $C(v_1) = C(v_2) \Rightarrow N(v_1) = N(v_2)$.
\end{theorem}

### 2022-03-21 concrete-semantics.tex
Edits for shepherd review and camera-ready

The \quark abstract machine of the previous section deliberately
ignores system-level concerns to focus on the semantics of
version-controlled state replication. In this section we present the
\quark \emph{distributed} machine that addresses the key system-level
concerns and provides a blueprint for a practical version-controlled
replicated state machine. We first reify the the one-to-one
correspondence between replicas and branches we assumed informally in
the previous section. This requires us to relax the assumption of the
state $\Delta$ being shared synchronously across all replicas. Next we
present an efficient method to track the version history with help
of vector clocks. Finally, we outline an algorithm for garbage
collecting older versions so that the version history need not grow
unboundedly.

@@

The state of the distributed machine is the triple $\delta = (B,N,H)$.
Components $B$ and $H$ are indexed by $\texttt{ReplicaId}$ to let us
denote their replica-local copies. For instance, replica $i$'s copy of
the \emph{head map} $H$ is given by $H\;i$, which we abbreviate to
$H_i$ for notational convenience. Like $H$ in
Fig.~\ref{fig:git-semantics}, $H_i$ maps each branch to the version
vector of its head. Note that two replicas may be out of sync w.r.t
the information in $H$ and $B$.  For instance, $H_i(b)$ may not be
equal to $H_j(b)$ if replicas $i$ and $j$ are out of sync. On a
replica $i$, \emph{branch map} $B_i$ gives the version vector
corresponding to a particular sequence number on a branch. For
instance, the version vector of the first version on branch $b$ is
given by $B_i(b,1)$ on replica $i$. The value map $N$ maps version
vectors to their values. We assume a single copy of $N$ to simplify
the presentation. Generalizing Fig.~\ref{fig:quark-semantics} to allow
for replica-local copies of $N$ is straightforward.

>>>>

The \quark abstract machine of the previous section deliberately
ignores system-level concerns to focus on the semantics of
version-controlled state replication. In this section we present the
\quark \emph{distributed} machine that addresses the key system-level
concerns and provides a blueprint for a practical version-controlled
replicated state machine. We first reify the one-to-one
correspondence between replicas and branches we assumed informally in
the previous section. This requires us to relax the assumption of the
state $\Delta$ being shared synchronously across all replicas. Next we
present an efficient method to track the version history with the help
of vector clocks. Finally, we outline an algorithm for garbage
collecting older versions so that the version history need not grow
unboundedly.

@@

The state of the distributed machine is the triple $\delta = (B,N,H)$.
Components $B$ and $H$ are indexed by $\texttt{ReplicaId}$ to let us
denote their replica-local copies. For instance, replica $i$'s copy of
the \emph{head map} $H$ is given by $(H\;i)$, which we abbreviate to
$H_i$ for notational convenience. Like $H$ in
Fig.~\ref{fig:git-semantics}, $H_i$ maps each branch to the version
vector of its head. Note that two replicas may be out of sync w.r.t
the information in $H$ and $B$.  For instance, $H_i(b)$ may not be
equal to $H_j(b)$ if replicas $i$ and $j$ are out of sync. On a
replica $i$, \emph{branch map} $B_i$ gives the version vector
corresponding to a particular sequence number on a branch. For
instance, the version vector of the first version on branch $b$ is
given by $B_i(b,1)$ on replica $i$. The value map $N$ maps version
vectors to their values. We assume a single copy of $N$ to simplify
the presentation. Generalizing Fig.~\ref{fig:quark-semantics} to allow
for replica-local copies of $N$ is straightforward.

### 2022-03-21 evaluation.tex
Edits for shepherd review and camera-ready

In this section we present an emperical evaluation of \quark on two
case studies. First is a collaborative document editing
application -- a common usecase addressed by several CRDT
proposals~\cite{rga, treedoc, crdts}. Second is a replicated key-value
store implemented using a mergeable Red-Black tree data structure.

@@

Our experiment setup consists of multiple collaborators simultaneously
editing a 10000+ line document obtained from the Canterbury
Corpus~\cite{canterbury}. Each user holds a replica of the document
and is assumed to be editing the document at the speed of 240
characters per minute or 1 character every 0.25s. At 6 characters per
word, this amounts to 40 words per minute, which is the average typing
speed of humans. Each edit is immediately persisted to the disk by
creating a new version in the backing store. Thus there are at least
as many versions of the document as there are edits. Such extensive
versioning may be considered excessive in practice and could be
disabled. Each user process runs a \quark thread that commits
user-generated document versions to the local branch, while
merging with the concurrent versions from the remote branches in the
background. Each merge is synchronized as described in previous
sections.

\quark's background merges however pose a new problem as they create
new versions on the local branch in the background while the user is
busy editing an older version. When the user attempts to write their
version of the document to the store, simply committing it would
effectively override the concurrent updates from other users obtained
via background merges. The solution, fortunately, is straightforward:
we \emph{merge} the user-submitted value with the (value of) the
latest version on the local branch to create a new version that
includes the updates from either direction. Since this merge is fully
confined to the local replica, it is guaranteed to not affect the LCA
of the local branch with respect to a remote branch. To the external
world, it appears as if the local branch has simply committed a new
version.
% The LCA for this merge is
% simply the last version on the local branch read by the user. This
% merge need not be synchronized as it doesn't alter the LCAs between
% any two branches. The merged value can then returned to the user as
% the result of the write.
Thus, \quark's \C{write} is a function of type $\texttt{Value}
\rightarrow \texttt{Value}$, where $\texttt{Value}$ is \C{doc} in the
current application.

\noindent\paragraph{Latency} To measure the impact of \quark runtime on user
writes, we measure the latency of the \C{write} operation, which
includes the time spent merging the user version with the current
version, and persisting the resultant version to the store. We conduct
the experiments on a three-node cluster of \C{i3.large} machines in
Amazon \C{us-west2} data center. Each user connects to one of the
machines, forks a new branch, and performs 1000 edits in succession,
saving the document after each edit. We progressively increase the
number of concurrent users editing the document from 3 to 200 and
measure the impact of the increased concurrency on write latency.
Fig.~\ref{fig:latency} shows the 10th, 50th (median), and 90th
percentile latency values. The median and 10th percentile latencies
remain more-or-less constant with a slight increase between 3 and 60
concurrent users. This demonstrates that \quark has negligible effect
on the per-operation latency, which is expected. The 90th percentile
latency, however, shows no clear pattern by the virtue of being more
susceptible to transient system conditions. Nonetheless, the maximum
value of 90th percentile latency measured (32ms) remains well below
the the time between consecutive edits (0.25s), making it hard to
perceive by a human user.

\begin{figure}[ht]
  \centering
    \includegraphics[scale=0.4]{Figures/monkey_latency}
\caption{Latency of writes to a shared document under \quark. }
\label{fig:latency}
\end{figure}

@@

\begin{figure}[ht]
  \centering
    \includegraphics[scale=0.4]{Figures/monkey_staleness}
\caption{Staleness increases as the number of concurrent editors
  increase.}
\label{fig:monkey-staleness}
  \vspace*{-0.2in}
\end{figure}

To understand the effect of \quark on
staleness, we quantify and measure it along with latency in the
experiment setup described above. Staleness is defined as the time
taken for a version committed on one replica to be merged into a
concurrent version on a remote replica. To measure staleness, we
annotate every version $v$ with the timestamp $t$ of capturing the
wall clock time of its creation. When $v$ is is merged into a remote
branch $b$ at a later time $t'$, the difference $t' - t$ denotes the
staleness of $v$ w.r.t the new version on $b$. One such staleness
measurement is recorded for every merge that ever happens during the
experiment. We compute 10th, 50th, and 90th percentiles of staleness
values thus obtained. Fig.~\ref{fig:monkey-staleness} shows the
results. As evident, staleness increases steadily with the increasing
number of concurrent replicas, which is expected considering that
merges are synchronized, and increasing the number of replicas reduces
the number of opportunities for a replica to merge. The increase is
roughly linear in the number of replicas as our implementation passes
the lock around in a round-robin fashion, thus increasing the lock
latency in proportion to the number of replicas. Nonetheless, the 90th
percentile staleness values remain low -- in the order of 100ms with
the number of replicas under 30, and in the order of 1s with number of
replicas under 60. While further optimizations might reduce staleness,
a non-trivial staleness overhead is inevitable in \quark due to our
use of synchronized merges to guarantee convergence.

@@

We repeat our experiments on a mergeable key-value store application
we implemented using a Red-Black tree MRDT. Latency results are shown
in Fig.~\ref{fig:rb-latency}, and staleness results in
Fig.~\ref{fig:rb-staleness}.

\begin{figure}[ht]
  \centering
    \includegraphics[scale=0.4]{Figures/rbmonkey_latency}
  \caption{Latency for Key-Value store operations under \quark}
\label{fig:rb-latency}
  \vspace*{-0.2in}
\end{figure}

\begin{figure}[ht]
  \centering
    \includegraphics[scale=0.4]{Figures/rbmonkey_staleness}
  \caption{Key-Value store staleness under \quark}
\label{fig:rb-staleness}
  \vspace*{-0.2in}
\end{figure}

Our experiments bring to the fore an inherent tradeoff among the
competing concerns of RDTs, namely (\rom{1}). The ease of programming
convergence, (\rom{2}) Latency, and (\rom{3}).  Staleness. While CRDTs
try to optimize for latency and staleness, they require a significant
amount of development and verification effort to be expended to ensure
convergence~\cite{kleppmann2017}. In contrast, \quark lets developers
derive convergent-by-construction RDTs from ordinary data data types
that are optimized for latency, but incur a staleness overhead that
delays the time to convergence.

>>>>

In this section we present an empirical evaluation of \quark on two
case studies. First is a collaborative document editing
application -- a common usecase addressed by several CRDT
proposals~\cite{rga, treedoc, crdts}. Second is a replicated key-value
store implemented using a mergeable Red-Black tree data structure.

@@

Our experiment setup consists of multiple collaborators simultaneously
editing a 10000+ line document obtained from the Canterbury
Corpus~\cite{canterbury}. Each user holds a replica of the document
and is assumed to be editing the document at the speed of 240
characters per minute or 1 character every 0.25s. At 6 characters per
word, this amounts to 40 words per minute, which is the average typing
speed of humans. Each edit is immediately persisted to the disk by
creating a new version in the backing store. Thus there are at least
as many versions of the document as there are edits. Such extensive
versioning may be considered excessive in practice and could be
disabled.

\noindent\paragraph{Latency} To measure the impact of \quark runtime on user
writes, we measure the latency of the \C{write} operation, which
includes the time spent merging the user version with the current
version, and persisting the resultant version to the store. We conduct
the experiments on a three-node cluster of \C{i3.large} machines
located in Amazon \C{us-west2} data center. Each user connects to one
of the machines, forks a new branch, and performs 1000 edits in
succession, saving the document after each edit. We progressively
increase the number of concurrent users editing the document from 3 to
60 and measure the impact of the increased concurrency on write
latency.  Fig.~\ref{fig:latency} shows the 10th, 50th (median), and
90th percentile latency values. The median and 10th percentile
latencies remain more-or-less constant with a slight increase between
3 and 60 concurrent users. This is expected considering that \quark
does not constrain the execution of user operations. The slight
increase, we confirmed, is due to the increased latency of concurrent
database writes. The 90th percentile latencies, however, show an
initial increase before adopting a pattern similar to median
latencies. We attribute this behavior to an idiosyncrasy of our
multi-threading implementation, which is more likely to affect
measurements at the extreme. In particular, we use OCaml v4.12
extended with Lightweight Threads (LWT) library, which offers
concurrency but not parallelism.  At lower number of replicas, the
merge thread's lock requests succeed more often, resulting in the
process spending more time merging than editing. We expect this
pattern to smoothen out in OCaml v5.0, which introduces true
multi-core parallelism. Notwithstanding this idiosyncrasy, the maximum
value of 90th percentile latency measured (32ms) remains well below
the the time between consecutive edits (0.25s), making it hard to
perceive by a human user.

\begin{figure}[ht]
  \centering
    \includegraphics[scale=0.4]{Figures/monkey_latency}
\caption{Latency of writes to a shared document under \quark. }
\label{fig:latency}
\vspace*{-0.2in}
\end{figure}

@@

\begin{figure}[ht]
  \centering
    \includegraphics[scale=0.4]{Figures/monkey_staleness}
\caption{Staleness increases as the number of concurrent editors
  increase.}
\label{fig:monkey-staleness}
% \vspace*{-0.2in}
\end{figure}

To understand the effect of \quark on
staleness, we quantify and measure it along with latency in the
experiment setup described above. Staleness is defined as the time
taken for a version committed on one replica to be merged into a
concurrent version on a remote replica. To measure staleness, we
annotate every version $v$ with the timestamp $t$ of the wall clock
time of its creation. When $v$ is is merged into a remote branch $b$
at a later time $t'$, the difference $t' - t$ denotes the staleness of
$v$ w.r.t the new version on $b$. One such staleness measurement is
recorded for every merge that ever happens during the experiment. We
compute 10th, 50th, and 90th percentiles of staleness values thus
obtained. Fig.~\ref{fig:monkey-staleness} shows the results. As
evident, staleness increases steadily with the increasing number of
concurrent replicas, which is expected considering that merges are
synchronized, and increasing the number of replicas reduces the number
of opportunities for a replica to merge. The increase is roughly
linear in the number of replicas as our implementation passes the lock
around in a round-robin fashion. Despite the steady increase, the 90th
percentile staleness values remain low -- in the order of 100ms with
the number of replicas under 30, and in the order of 1s with number of
replicas under 60. While further optimizations might reduce staleness,
a non-trivial staleness overhead is inevitable in \quark due to our
use of synchronized merges to guarantee convergence.

@@

Our second case study involves a mergeable key-value store, which,
unlike a text editor, is expected to work in the background supporting
applications such as e-Commerce. The store is mergeable in the sense
that it can merge conflicting values assigned to equal keys using
value-specific merge functions. For instance, an e-Commerce
application might store shopping carts as values, in which case the
merge function on shopping carts is used to merge concurrent
conflicting updates to a user's shopping cart. The key-value store is
implemented using a Red-Black tree MRDT~\cite{mrdt}, which is a
standard Red-Black tree implementation extended with a merge function.
The merge function implements set merge semantics
(Sec.~\ref{sec:motivation}) for concurrent insertions and deletions of
keys, and defers to the value merge function for concurrent updates
against a key. Concretely, \C{RBTree} is an OCaml functor
parameterized on the \C{Key} and \C{Value} modules, where the latter
is required to provide a \C{merge} function. We use integer keys and
mergeable counter values in our experiments.

% \begin{figure}[ht]
%   \centering
%     \includegraphics[scale=0.4]{Figures/rbmonkey_latency}
%   \caption{Latency of operations on a key-value store implemented
%   using a Red-Black tree MRDT.}
% \label{fig:rb-latency}
%   \vspace*{-0.2in}
% \end{figure}

\begin{figure}[ht]
  \centering
    \includegraphics[scale=0.4]{Figures/rbmonkey_staleness}
  \caption{Staleness of key-value store merge operations.}
\label{fig:rb-staleness}
% \vspace*{-0.2in}
\end{figure}

We use the same experimental setup as before -- an increasing number
of concurrent processes perform a random operation on the local
version of the tree while merging with the remote versions in the
background.  Each process forks off a branch with an initial version
of an \C{RBTree} of size 1000, performs an insertion, updatation, or
deletion with probabilities of 0.5, 0.25, and 0.25. A new version of
the tree is committed after each operation. Each process performs 1000
operations with an inter-operation delay of 100ms. We measure the
latency of each tree operation and the staleness of each merge.
Latency measurements follow the same pattern as those for
collaborative editing (Fig.~\ref{fig:latency}), and have been elided.
The median latency remains around 50ms for most of the experiment.
Staleness measurements are plotted in Fig.~\ref{fig:rb-staleness}. As
before, staleness increases steadily in proportion to the increasing
number of concurrent processes. The 90th percentile staleness exceeds
1s for 15 concurrent processes, and 7s for 54 concurrent processes.
Unlike the collaborative applications, however, the number of replicas
in a conventional web application tends to be low, which could limit
staleness to a tolerable range. For instance, in the e-Commerce
application described above, 3 replicas of the shopping cart can be
kept in sync with each other with a 90th percentile lag (staleness) of
30ms.

Our experiments bring to the fore an inherent tradeoff among the
competing concerns of Convergent RDTs, namely (\rom{1}). The ease of
programming, (\rom{2}) Latency, and (\rom{3}).  Staleness. While CRDTs
try to optimize for latency and staleness, they require a significant
amount of development and verification effort to be expended to ensure
convergence. In contrast, \quark lets developers derive
convergent-by-construction MRDTs from ordinary data types that are
optimized for latency, but incur a non-trivial staleness overhead that
delays the time to convergence.

### 2022-03-21 implementation.tex
Edits for shepherd review and camera-ready

The implementation of \quark is a straightforward realization of the
\quark distributed machine described in Sec.~\ref{sec:concrete-sem}.
We manifest each component of the state, namely the branch map $B$,
the value map $N$, and the head map $H$, as a column family (i.e., a
table) in Scylla. The synchronization needed to linearize merges is
implemented with help of Scylla's support for conditional updates (CAS
operations) and expiring columns. The total order among merges is
enforced with help of \C{Quorum} reads and writes. Each user process
is assigned its own replica of the MRDT.  Version vectors are realized
as associative lists and stored in Scylla as blobs.

One key difference between the formalization and the implementation is
in the treatment of MRDT values. Formalization assumes values to be
atomic with no sharing in between them. In practice, however, an MRDT
could be a linked data structure such as a binary tree, and two
such values could share a significant amount of internal structure.
Consequently, the size of the \emph{diff} between two consecutive
versions of a value could be asymptotically less then the size of the
data structure itself, in which case it unreasonable to transfer the
entire data structure over the network. To facilitate the efficient
computation of diffs between versions of data structures, we implement
a \emph{content-addressable store} as a key-value table in Scylla
where key is simply the SHA256 hash of the value. A linked data
structure is stored as a collection of nodes, where each node links to
the other by referring to its hash. The diff between two consecutive
versions of a data structure would simply manifest as new entries in
the content-addressable store reachable from the root of the new
version. The new entries, being new data, are automatically replicated
by Scylla, thus letting us reconstruct the new version at a remote
location. The root hash of the new version is obtained from the value
map $N$, which now maps version vectors to the \emph{hashes} of
values.

>>>>

The implementation of \quark largely follows the design of \quark
distributed machine described in Sec.~\ref{sec:concrete-sem}.  We
manifest each component of the state, namely the branch map $B$, the
value map $N$, and the head map $H$, as a column family (i.e., a
table) in Scylla. The synchronization needed to linearize merges is
implemented with help of Scylla's support for conditional updates (CAS
operations) and expiring columns. The total order among merges is
enforced with help of \C{Quorum} reads and writes. Each user process
is assigned its own branch containing a replica of the MRDT.  Version
vectors are realized as associative lists and stored in Scylla as
blobs.

The implementation however differs from the formalization in two
significant ways. First is in the treatment of MRDT values.
Formalization assumes values to be atomic with no sharing in between
them. In practice, however, an MRDT could be a linked data structure
such as a binary tree, and two such values could share a significant
amount of internal structure.  Consequently, the size of the
\emph{diff} between two consecutive versions of a value could be
asymptotically less then the size of the data structure itself, in
which case it unreasonable to transfer the entire data structure over
the network. To facilitate the efficient computation of diffs between
versions of data structures, we implement a \emph{content-addressable
store} as a key-value table in Scylla where key is simply the SHA256
hash of the value. A linked data structure is stored as a collection
of nodes, where each node links to the other by referring to its hash.
The diff between two consecutive versions of a data structure would
simply manifest as new entries in the content-addressable store
reachable from the root of the new version. The new entries, being new
data, are automatically replicated by Scylla, thus letting us
reconstruct the new version at a remote location. The root hash of the
new version is obtained from the value map $N$, which now maps version
vectors to the \emph{hashes} of values.

The second significant difference between the implementation and the
formalization is the timing of merges. Operational Semantics in
Sec.~\ref{sec:abstract-sem} and Sec.~\ref{sec:concrete-sem} interleave
commits and merges such that only one of them executes at any given
time. Since commits are initiated by the user in practice,
interleaving them with strongly consistent merges increases their
latency as perceived by the user. For the user-perceived latency to
remain unaffected, it is important that a replica be always available
to execute user requests. \quark ensures this by handling user
requests in a foreground thread that is always allowed to commit to
the local branch. Merges, on the other hand, are relegated to the
background. A background thread constantly scans the remote branches
for new versions, and if there are any, merges them into the latest
version on the local branch after checking the necessary preconditions
(Sec.~\ref{sec:abstract-sem}).

\quark's background merges however pose a new problem as they create
new versions on the local branch in the background while a user
operation is manipulating an older version in the foreground. When
the user attempts to write their version to the store, simply
committing it would effectively override the concurrent updates from
other users obtained via background merges. The solution, fortunately,
is straightforward: \quark \emph{merges} the user-submitted version
with the latest version on the local branch to create a new version
that includes the updates from either direction. Since this merge is
fully confined to the local replica, it is guaranteed to not affect
the LCA of the local branch in relation to any remote branch. To the
external world, it appears as if the local branch has simply committed
a new version that was derived from the older version. Since the
commit operation now entails a merge, the merged version has to be
returned to the user as the result of the commit.  Consequently the
\C{write} operation exposed by \quark is a function of type
$\texttt{Value} \rightarrow \texttt{Value}$, where $\texttt{Value}$ is
any MRDT.
